*R Script for Bat Mass Data Analysis (Group-G)*
Group-G Members: [Andrew Batmunkh] [Yunjung Jo] [Victoria Cordova-Morote]
Date: [2024-12-07]

Description: This script analyzes bat mass data to evaluate the effects of sex, pregnancy status, and disease group (pre- vs. post-invasion) using statistical modeling and visualization. 
Key tasks include:
 1. Preparing data by creating categorical variables and combinations.
 2. Fitting Gamma and Normal distributions using Maximum Likelihood Estimation (MLE)     and comparing model fits with Akaike Information Criterion (AIC).
 3. Computing profile likelihood-based confidence intervals for Gamma distribution       parameters.
 4. Fitting a Generalized Linear Mixed Model (GLMM) to incorporate random effects for     site-level grouping.
 5. Visualizing observed data, fitted distributions, and model predictions across        predictor groups.
 6. Combining results into a comprehensive data frame for further analysis.


```{r}
require(tidyverse)
require(fitdistrplus)
require(tidyverse)
require(deSolve)
library(lme4)
```
#read the data onto R as "data
```{r}
data <- read_csv("/Users/andrewbatmunkh/Desktop/eeb313/Big.csv")
head(data)
```

#This code prepares data, fits Gamma and Normal distributions using MLE, compares model fits with AIC, computes confidence intervals for Gamma parameters, and combines results for further analysis.
```{r}
# Data preparation
data_new <- data |> 
  mutate(
    disease_group = ifelse(disease_time_step == "pre-invasion", 
                           "Pre-Invasion", 
                           "Post-Invasion"),
    pregnancy_status = ifelse(repstat == "pregnant", 
                              "Pregnant", 
                              "Not Pregnant")
  )

# Convert relevant columns to character type
data_new$age <- as.character(data_new$age)
data_new$sex <- as.character(data_new$sex)
data_new$disease_group <- as.character(data_new$disease_group)
data_new$pregnancy_status <- as.character(data_new$pregnancy_status)

# Create all combinations of age, sex, disease group, and pregnancy status
combos <- expand.grid(
  age = as.character(unique(data_new$age)),
  sex = as.character(unique(data_new$sex)),
  disease_group = as.character(unique(data_new$disease_group)),
  status = as.character(unique(data_new$pregnancy_status))
)

# Generate combinations of predictor variables and filter invalid combinations
combos |> 
  subset(!(sex == "male" & status == "Pregnant")) |> 
  subset(!(sex == "female" & status == "Not Pregnant")) -> combos

rownames(combos) <- NULL  

# Function to find MLE and fit distributions
MLE_finder <- function(age, sex, group, status) {
  
  # Filter data for the specified combination
  # This will ensure only data relevant to the current group is analyzed.
  data_new_new <- data_new[data_new$age == paste(as.character(age)) &
                           data_new$sex == paste(as.character(sex)) &
                           data_new$disease_group == paste(as.character(group)) &
                           data_new$pregnancy_status == paste(as.character(status)), ]
  
  # Check if there is enough data for fitting on Gamma and Normal distributions
  if (length(data_new_new$mass) < 2) {
    print("Not enough data!")
    return(NA)
  } else {
    
    # Fit Gamma and Normal distributions
    best_fit_Gamma <- fitdist(data_new$mass, distr = "gamma", method = "mle")
    best_fit_Normal <- fitdist(data_new$mass, distr = "norm", method = "mle")
    
    # Compare AIC values to determine the better fit
    betterFit <- best_fit_Gamma$aic < best_fit_Normal$aic - 2
    print(betterFit)
    
    #  Gamma distribution fit
    cbind(
      prob = dgamma(seq(min(data_new_new$mass), max(data_new_new$mass), length = 100), 
                    shape = best_fit_Gamma$estimate[1],
                    rate = best_fit_Gamma$estimate[2]),
      mass = seq(min(data_new_new$mass), max(data_new_new$mass), length = 100)
    ) -> fitGamma
    
    # Normal distribution fit
    cbind(
      prob = dnorm(seq(min(data_new_new$mass), max(data_new_new$mass), length = 100), 
                   mean = best_fit_Normal$estimate[1],
                   sd = best_fit_Normal$estimate[2]),
      mass = seq(min(data_new_new$mass), max(data_new_new$mass), length = 100)
    ) -> fitNormal
    
    # Plot histogram with distribution fits
    data_new_new |> 
      ggplot(aes(x = mass)) +
      geom_rug() +
      geom_histogram(aes(y = ..density..)) +
      geom_line(data = fitGamma, aes(x = mass, y = prob), color = "red") +
      labs(title = paste0(group, " distribution of ", age, " ", sex, " masses")) -> p
    
    print(p)
    
    return(best_fit_Gamma)
  }
}

# Apply MLE_finder to all combinations
results <- NULL
for (i in 1:nrow(combos)) {
  MLE_finder(
    age = combos[i, "age"], 
    sex = combos[i, "sex"], 
    group = combos[i, "disease_group"],
    status = combos[i, "status"]
  ) -> results[[i]]
}

# Function to find confidence intervals for Gamma distribution parameters
CI_finder <- function(age, sex, group, status) {
  
  # Filter data for the specified combination
  data_new_new <- data_new[data_new$age == paste(as.character(age)) &
                           data_new$sex == paste(as.character(sex)) &
                           data_new$disease_group == paste(as.character(group)) &
                           data_new$pregnancy_status == paste(as.character(status)), ]
  
  # Check if there is enough data for CI calculation
  if (length(data_new_new$mass) < 2) {
    print("Not enough data!")
    return(NA)
  } else {
    
    # Reuse MLE_finder for parameter estimates
    results <- MLE_finder(age, sex, group, status)
    maxLL <- sum(dgamma(data_new_new$mass,
                        shape = results$estimate[1], 
                        rate = results$estimate[2], log = TRUE))
    
    shape <- results$estimate[1]
    rate <- results$estimate[2]
    
    # Confidence intervals for shape parameter
    LL_rate <- NULL
    Shapes <- seq(shape / 2, shape * 2, length = 1000)
    for (i in 1:length(Shapes)) {
      sum(dgamma(data_new_new$mass,
                 shape = Shapes[i], 
                 rate = rate, log = TRUE)) -> LL_rate[i]
    }
    data.frame(LL = LL_rate, shape = Shapes, rate = rate) -> CI_shape
    
    # Confidence intervals for rate parameter
    LL_shape <- NULL
    Rates <- seq(rate / 2, rate * 2, length = 1000)
    for (i in 1:length(Rates)) {
      sum(dgamma(data_new_new$mass,
                 shape = shape, 
                 rate = Rates[i], log = TRUE)) -> LL_shape[i]
    }
    data.frame(LL = LL_shape, shape = shape, rate = Rates) -> CI_rate
    
    return(
      cbind(
        rbind(CI_shape, CI_rate), 
        age = as.character(age), 
        sex = as.character(sex), 
        disease_group = as.character(group), 
        status = as.character(status)
      )
    )
  }
}

# Apply CI_finder to all combinations
results_pt2 <- NULL
for (i in 1:nrow(combos)) {
  CI_finder(
    age = combos[i, "age"], 
    sex = combos[i, "sex"], 
    group = combos[i, "disease_group"],
    status = combos[i, "status"]
  ) -> results_pt2[[i]]
}

# Combine results into a single data frame for analysis
as.data.frame(do.call(rbind, results_pt2)) -> datAnalyze

```
# Subsets post-invasion adult male mass data, fits Gamma and Normal distributions, and visualizes the observed data with fitted density curves.
```{r}
# Subset the data for post-invasion adult males
data_subset <- data_new[data_new$disease_group == "Post-Invasion" &
                          data_new$age == "adult" &
                          data_new$sex == "male", ]

# Fit a normal distribution to the data
fit_normal <- fitdist(data_subset$mass, distr = "norm", method = "mle")

# Fit a gamma distribution to the data
fit_gamma <- fitdist(data_subset$mass, distr = "gamma", method = "mle")

#Generate a density function based on the fitted normal distribution
fit_normal_data <- data.frame(
  mass = seq(min(data_subset$mass), max(data_subset$mass), length.out = 100),
  density = dnorm(seq(min(data_subset$mass), max(data_subset$mass), length.out = 100),
                  mean = fit_normal$estimate["mean"],
                  sd = fit_normal$estimate["sd"])
)

#Generate a density function based on the fitted gamma distribution
fit_gamma_data <- data.frame(
  mass = seq(min(data_subset$mass), max(data_subset$mass), length.out = 100),
  density = dgamma(seq(min(data_subset$mass), max(data_subset$mass), length.out = 100),
                   shape = fit_gamma$estimate["shape"],
                   rate = fit_gamma$estimate["rate"])
)

#Plot the histogram of mass data with fitted density curves
ggplot(data_subset, aes(x = mass)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, fill = "grey", alpha = 0.7, color = "black") +
  geom_rug() + # Add rug plot
  geom_line(data = fit_normal_data, aes(x = mass, y = density), color = "red", size = 1) +
  geom_line(data = fit_gamma_data, aes(x = mass, y = density), color = "blue", size = 1) +
  labs(
    title = "Post-Invasion Distribution of Adult Male Masses",
    x = "Mass",
    y = "Density"
  ) 
```

## Confidence Interval Analysis of the bat data and Visualization of Gamma Distribution Parameters
```{r}
# Ensure numeric type for Gamma distribution parameters
datAnalyze$shape <- as.numeric(datAnalyze$shape)
datAnalyze$rate <- as.numeric(datAnalyze$rate)

# Identify Maximum Likelihood Estimates (MLEs) for each group
datAnalyze |> 
  group_by(age, sex, disease_group, status) |> 
  mutate(MLE = case_when(
    LL == max(LL) ~ "yes",
    LL != max(LL) ~ "no"
  )) |> 
  subset(MLE == "yes") |> 
  mutate(mean = shape / rate) |> 
  ggplot(aes(x = sex, y = mean)) +
  geom_point()  # Plot the mean (No CIs)

# Extract only the MLE rows for further analysis
datAnalyze |> 
  group_by(age, sex, disease_group, status) |> 
  mutate(MLE = case_when(
    LL == max(LL) ~ "yes",
    LL != max(LL) ~ "no"
  )) |> 
  subset(MLE == "yes") -> MLEs

# Initialize storage for confidence intervals
CIs_shape <- NULL
CIs_rate <- NULL

# Loop through MLEs to calculate confidence intervals for shape and rate
for (j in 1:nrow(MLEs)) {
  # Plot profile likelihood for shape
  print(
    datAnalyze |> 
      subset(age == MLEs$age[j] & sex == MLEs$sex[j] &
             disease_group == MLEs$disease_group[j] & 
             status == MLEs$status[j]) |> 
      subset(rate == MLEs$rate[j]) |> 
      ggplot(aes(x = shape, y = LL)) + 
      geom_line()
      # geom_hline(yintercept = MLEs$LL[j] - 1.92, color = "red")
  )

  # Plot profile likelihood for rate
  print(
    datAnalyze |> 
      subset(age == MLEs$age[j] & sex == MLEs$sex[j] & 
             disease_group == MLEs$disease_group[j] & 
             status == MLEs$status[j]) |> 
      subset(shape == MLEs$shape[j]) |> 
      ggplot(aes(x = rate, y = LL)) + 
      geom_line()
  )

  # Confidence intervals for shape
  datAnalyze |> 
    subset(age == MLEs$age[j] & sex == MLEs$sex[j] & 
           disease_group == MLEs$disease_group[j] & 
           status == MLEs$status[j]) |> 
    subset(rate == MLEs$rate[j]) |> 
    summarise(upper_shape = max(shape), lower_shape = min(shape)) -> CIs_shape[[j]]

  # Confidence intervals for rate
  datAnalyze |> 
    subset(age == MLEs$age[j] & sex == MLEs$sex[j] & 
           disease_group == MLEs$disease_group[j] & 
           status == MLEs$status[j]) |> 
    subset(shape == MLEs$shape[j]) |> 
    summarise(upper_rate = max(rate), lower_rate = min(rate)) -> CIs_rate[[j]]
}

# Combine MLEs with confidence intervals into a single data frame
as.data.frame(cbind(MLEs, do.call(rbind, CIs_shape), do.call(rbind, CIs_rate)))[-8] -> ALLCIs

# Note: shape/rate is the mean of the Gamma distribution, and shape/rate^2 is its variance

# Plot mean of Gamma distribution by various groups
ALLCIs |> 
  ggplot(aes(x = disease_group, y = shape / rate)) + geom_point()
ALLCIs |> 
  ggplot(aes(x = sex, y = shape / rate)) + geom_point()
ALLCIs |> 
  ggplot(aes(x = age, y = shape / rate)) + geom_point()
ALLCIs |> 
  ggplot(aes(x = status, y = shape / rate)) + geom_point()

# Plot variance of Gamma distribution by various groups
ALLCIs |> 
  ggplot(aes(x = disease_group, y = shape / rate^2)) + geom_point()
ALLCIs |> 
  ggplot(aes(x = sex, y = shape / rate^2)) + geom_point()
ALLCIs |> 
  ggplot(aes(x = age, y = shape / rate^2)) + geom_point()
ALLCIs |> 
  ggplot(aes(x = status, y = shape / rate^2)) + geom_point()

# Plot shape parameter with confidence intervals
ALLCIs |> 
  ggplot(aes(
    x = sex, y = shape,
    ymin = lower_shape, ymax = upper_shape,
    group = interaction(disease_group, age, status)
  )) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(width = 0.1, position = position_dodge(width = 0.5))

# Plot scale parameter with confidence intervals
ALLCIs |> 
  ggplot(aes(
    x = sex, y = 1 / rate,
    ymin = 1 / lower_rate, ymax = 1 / upper_rate,
    group = interaction(disease_group, age, status)
  )) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(width = 0.1, position = position_dodge(width = 0.5))

# Plot mean with confidence intervals
ALLCIs |> 
  mutate(
    scale = 1 / rate,
    lower_scale = 1 / upper_rate,
    upper_scale = 1 / lower_rate
  ) |> 
  ggplot(aes(
    x = sex, y = shape * scale,
    ymin = lower_shape * lower_scale,
    ymax = upper_shape * upper_scale,
    group = interaction(disease_group, age, status)
  )) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(width = 0.1, position = position_dodge(width = 0.5))

# Plot variance with confidence intervals
ALLCIs |> 
  mutate(
    scale = 1 / rate,
    lower_scale = 1 / upper_rate,
    upper_scale = 1 / lower_rate
  ) |> 
  ggplot(aes(
    x = sex, y = shape * scale^2,
    ymin = lower_shape * lower_scale^2,
    ymax = upper_shape * upper_scale^2,
    group = interaction(disease_group, age, status)
  )) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(width = 0.1, position = position_dodge(width = 0.5))

```

# This code fits a GLMM to model bat mass as a function of sex, pregnancy status, and disease group, including a random intercept for site_mask. 
```{r}
glmer(
  mass ~ sex * pregnancy_status * disease_group + (1 | state), 
  data = data_new, 
  family = Gamma(link = "log")
) -> model

# Display the model summary
summary(model)
```

# LMM of bat mass as a function of sex, pregnancy status, and disease group, including a random intercept for site_mask. This is used to compare the LMM and GLMM models
```{r}
lmer(
  mass ~ sex * pregnancy_status * disease_group + (1 | site_mask), 
  data = data_new
) -> lmm_model

# Display the model summary
summary(lmm_model)

```

# This code runs the visualization of the predicted values with and without random effects across groups, faceted by sex and pregnancy status.
```{r}
# Predict values from the model, both with and without random effects
as.data.frame(data_new)[, c("mass", "pregnancy_status",  "disease_group", "sex", "site_mask")] |> 
  mutate(fit.m = predict(model, re.form = NA), # does not include random effects
         fit.c = predict(model, re.form = NULL) # includes random effects
         ) ->
  predicted_values

# Visualization of predicted values
predicted_values |>
  ggplot(aes(x = factor(disease_group, level = c("Pre-Invasion", "Post-Invasion")), 
             group = site_mask)) +
  geom_line(aes(y = fit.c,
                color = site_mask, 
                group = site_mask), 
            size = 2) +
  geom_line(aes(y = fit.m, group = ""), 
            color = "black", size = 3) +
  theme_classic() +
  facet_grid(sex~pregnancy_status) +
  theme(legend.position = "none" )
```


